\chapter{Definite (Un)Assignment}
\label{c_definite_assignment}

The Whiley programming language requires that variables are known at
\gls{compile_time} to be {\em definitely assigned} (i.e. that they are
defined before used) and, similarly, that \lstinline{final} variables
are {\em definitely unassigned} at the point of assignment (i.e. that
they are assigned at most once).  A conservative approach is taken to
determining whether or not these requirements hold true.  This ensures
the language can be compiled efficiently, but also means that some
provably safe programs are not valid Whiley programs.  In this
chapter, we specify the process by which definite assignment and
unassignment are determined.  The mechanism underpinning this is a
data-flow analysis over the control-flow graph of a function to
determine the assignment status of all local variables.

\section{Overview}

The following illustrates a simple function which will be rejected by
the compiler because it cannot determine definite assignment for all
variables.  The function is said to {\em fail definite assignment}:

\begin{lstlisting}
function f(int x) -> (int r):
   int y
   //
   if x < 0:
       y = 1
   //
   return x + y
\end{lstlisting}

In the above program, variable \lstinline{y} is {\em not} definitely
assigned before its use in the \lstinline{return} statement.  This is
because there is an \gls{execution_path} through the function which
reaches the \lstinline{return} statement and on which variable
\lstinline{y} is not defined (see Figure~\ref{f:c8_cfg_1}).  In fact,
there are {\em two} possible execution paths through this function,
but variable \lstinline{y} is only defined on one of them.  Observe
that, since it is a parameter, variable \lstinline{x} is automatically
considered to have been defined on entry to the function.

In contrast, the following illustrates a simple function which will be
rejected by the compiler because it cannot determine a
\lstinline{final} variable is definitely unassigned at the point of
assignment:

\begin{lstlisting}
function g(int x) -> (int r):
   //
   final int y = 0
   //
   if x < 0:
       y = 1
   //
   return y
\end{lstlisting}

In the above program, the \lstinline{final} variable \lstinline{y} is
{\em not} definitely unassigned at the assignment on line 5.  In other
words, there is an \gls{execution_path} through the function on which
variable \lstinline{y} is assigned more than once.  Such a path
violates the intention of the \lstinline{final} modifier, which
dictates that a variable should be assigned at most once.

\begin{figure}[!t]
\centering
  \begin{tikzpicture}[
    graynode/.style={
      rectangle, 
      draw=gray!100, 
      fill=gray!20.5, 
      very thick, 
      minimum size=5mm,
      minimum height=1cm,
      minimum width=4cm,
      rounded corners=0.25cm
    }
    ]
    \node[graynode] (l1) at (0,4cm) {\lstinline{x < 0}};
    \node at (1,3.25) {\tiny${\tt true}$};
    \node at (-1,3.25) {\tiny${\tt false}$};
    \node[graynode] (l2) at (2.5cm,2cm) {\lstinline{y = 1}};
    \node[graynode] (l3) at (0,0) {\lstinline{return x + y}};
    \draw[thick,->] (l1)++(0.5,-0.5) -- (0.5,3) -- (2.5,3) -- (l2);
    \draw[thick,->] (l2) -- (2.5,1) -- (0.5,1) -- (0.5,0.5);
    \draw[ultra thick,->] (-0.5,3.5) -- (-0.5,0.5);
  \end{tikzpicture}
\caption{Control-Flow Graph for function \lstinline{f()}.  On the bold path, \lstinline{y} is undefined.}
\label{f:c8_cfg_1}
\end{figure}

\subsection{Loops}

The treatment of loops with respect to definite (un)assignment
warrants special attention.  Recall the mechanism for determining
definite (un)assignment is conservative.  In the context of loops, it
simply assumes {\em every loop can be executed zero or more times}
(i.e. even if this is not correct).  From the perspective of definite
assignment, this implies that a variable assignment within a loop {\em
  may not occur}.  From the perspective of definite unassignment, this
implies that a variable assignment within a loop {\em may occur}. The
following illustrates:

\begin{lstlisting}
function f(int x) -> int:
   int y
   //
   while x < 0:
       y = 1
       x = x + 1
   //
   return x + y
\end{lstlisting}

The above function fails definite assignment because variable
\lstinline{y} is not defined when zero iterations of the loop are
executed (e.g. when \lstinline{x==0} on entry).  In contrast, the
following illustrates a function which fails definite unassignment:

\begin{lstlisting}
function g(int x) -> int:
   final int y = 0
   //
   while x < 0:
       y = 1
       x = x + 1
   //
   return x + y
\end{lstlisting}

The above function fails definite unassignment because variable
\lstinline{y} is defined more than once when one or more iterations of
the loop are executed (e.g. when \lstinline{x<0} on entry).  To
illustrate the conservative nature of definite assignment, consider
this variation:

\begin{lstlisting}
function ten() -> int:
   int x = 0
   int y
   //
   while x < 10:
       y = 1
       x = x + 1
   //
   return y
\end{lstlisting}

Here, it can be shown that \lstinline{y==1} must hold when the
\lstinline{return} statement is reached.  Nevertheless, this function
fails definite assignment because of the assumption that the loop
executes {\em zero} or more iterations.

\subsection{Infeasible Paths}

Functions and methods may contain \glslink{infeasible_path}{infeasible
  paths} which are valid execution paths that, in practice, cannot be
executed.  The mechanism for checking definite (un)assignment assumes
for simplicity that any valid path can be executed.  This means that
some programs will fail definite assignment, even though they can be
shown as safe.  The following illustrates such a program:

\begin{lstlisting}
function abs(int x) -> int:
    int y
    //
    if x >= 0:
        y = x
    //
    if x < 0:
        y = -x
    //
    return y
\end{lstlisting}

This function contains four valid execution paths which can be denoted
by \lstinline{ff}, \lstinline{tf}, \lstinline{ft}, \lstinline{tt}
where, for example, \lstinline{tf} represents the path where the first
condition evaluates to \lstinline{true} and the second to
\lstinline{false}.  However, it is easy to see that execution paths
\lstinline{ff} and \lstinline{tt} are infeasible.  Furthermore, that
on the other two paths, \lstinline{tf} and \lstinline{ft}, variable
\lstinline{y} is definitely assigned at the \lstinline{return}
statement.  Despite this, the above function fails definite assignment
because the mechanism considers all valid paths whilst ignoring
infeasible execution paths.

\subsection{Partial Assignments}

A variable will never be considered definitely assigned after the
application of one or more {\em partial assignments}.  In contrast, a
variable is no longer considered definitely unassigned in such case.
The following illustrates a program which fails definite assignment
even though it can be shown as safe:

\begin{lstlisting}
type Point is {int x, int y}

function Point(int x, int y) -> Point:
    Point p
    p.x = x
    p.y = y
    return p
\end{lstlisting}

In the above program, the variable \lstinline{p} can be shown as definitely assigned at the \lstinline{return} statement.  Nevertheless, the conservative mechanism for checking definite assignment will reject this program because variable \lstinline{p} is initialised via partial assignments.  

\section{Description}

Definite assignment is defined over the \gls{control_flow_graph} of a code block, such as a \lstinline{requires} or \lstinline{ensures} clause, or the body of a function or method.  Appendix~\S\ref{c_cfg} details the process for constructing a control-flow graph from a block of code.  

Definite assignment considers the set of all {\em valid paths} in a \gls{control_flow_graph}.  A valid path is a path through the graph starting from its root.  Every vertex in the graph is associated with a set of variables which are {\em defined} at that vertex, as well as those which are {\em used} at that vertex.  A variable $x$ is said to be {\em definitely assigned} on entry to a vertex $v$ if, for every valid path which includes $v$, some ancestor vertex $u$ exists on which $x$ is defined.

\subsection{Definitions and Uses}

A \gls{variable_definition} occurs at a vertex $v$ which contains a direct assignment (recall~\S\ref{c_stmt_assign}), or an initialisation of (recall~\S\ref{c_stmts_var_decl}), one or more variables.  Assignments to fields, array elements or dereferenced references are not direct and do not define variables.  A vertex may define multiple variables if it corresponds to a multiple assignment which directly assigns those variables.  Finally, the parameters of a function, method or invariant block are treated as being defined on entry. 

A \gls{variable_use} occurs at a vertex $v$ which directly refers to that variable in some expression other than an \lstinline{LVal}.  However, referring to a variable indirectly through a dereference expression does not constitute a variable use.
