\chapter{Errors and Warnings}

When the Whiley compiler encounters an invalid program it will report an {\em error}.  In contrast, when it encounters something undesirable in an otherwise valid program, it may report a {\em warning}.  This chapters details the complete list of error messages and warnings which can be reported for a Whiley program.

\section{Overview}

\section{Parse Errors}

\section{Declarations}

Declarations are top-level entities in a \gls{source_file}, and their syntax is defined in \S\ref{c_source_files}.

\subsection{``Cyclic Constant Declaration'' (E301)}

A {\em cyclic constant declaration} occurs when a \gls{constant_declaration} refers to itself, either {\em directly} or {\em indirectly}.  This is an error because constants must be evaluated at \gls{compile_time}.

\paragraph{Example.}  The following illustrates several cyclic constant declarations:

\begin{lstlisting}
constant const1 is 1 + const1

constant const2 is 1 + const3
constant const3 is 1 + const2
\end{lstlisting}
Here, all three constant declarations are cyclic.  The declaration for \lstinline{const1} has a {\em direct} cycle, because its definition refers to itself.  The declaration for \lstinline{const2} has an indirect cycle, because its definition refers to \lstinline{const3} which, in turn, refers back to \lstinline{const2}.

\subsection{``Reference Not Permitted in Function'' (E302)}

A {\em reference not permitted in function} error occurs when an attempt is made to declare or use a variable of reference type in a function (as opposed to a method).  Functions in Whiley must be free from \glspl{side_effect} --- i.e. they must be {\em pure}.  This, the potential side-effects made possible through the use of references is not permitted.

\paragraph{Example.}  The following illustrates a very simple example:

\begin{lstlisting}
function f(&int x) -> (&int r):
   return x
\end{lstlisting}

Here, function \lstinline{f()} accepts a parameter \lstinline{x} of reference type \lstinline{&int}, which is not permitted.  In this case the function does not, in fact, exhibit any side-effects; nevertheless, the function will currently be rejected.

\subsection{``Reference Operation Not Permitted in Function'' (E303)}

A {\em reference operation not permitted in function} error occurs when an attempt is made to operate on a variable of reference type in a function (as opposed to a method).  Functions in Whiley must be free from \glspl{side_effect} --- i.e. they must be {\em pure}.  This, the potential side-effects made possible through the use of references is not permitted.

\paragraph{Example.}  The following illustrates a very simple example:

\begin{lstlisting}
function f(int x) -> (int r):
   int y = x
   return *(&y)
\end{lstlisting}

Here, function \lstinline{f()} obtains a reference to local variable \lstinline{y} and then immediately dereferences it, neither of which is permitted.  In this case the function does not, in fact, exhibit any side-effects; nevertheless, the function will currently be rejected.

\subsection{``Method Invocation Not Permitted In Function'' (E304)}

A {\em method invocation not permitted in function} error occurs when an attempt is made to call a method from a function (as opposed to another method).  Functions in Whiley must be free from \glspl{side_effect} --- i.e. they must be {\em pure}.  This, the potential side-effects made possible through the method call are not permitted.

\paragraph{Example.}  The following illustrates a very simple example:

\begin{lstlisting}
method g(int x) -> (int y):
   return x

function f(int x) -> (int r):
   return g(x)
\end{lstlisting}

Here, function \lstinline{f()} accepts a parameter \lstinline{x} and passes it through a call to method \lstinline{g()}.  In this case method \lstinline{g()} does not, in fact, exhibit any side-effects; nevertheless, the method call will be rejected.

\subsection{``Return from Void'' (E305)}

A {\em return from void} error occurs when an attempt is made to return a value from a function or method either explicitly declared as returning \lstinline{void} or where no return type is given (and, hence, is implicitly declared as returning \lstinline{void}).

\paragraph{Example.}  The following illustrates two simple examples:

\begin{lstlisting}
method g(int x):
   return x+1

function f(int x) -> void:
   return x
\end{lstlisting}

Here, method \lstinline{g()} is {\em implicitly} declared as returning \lstinline{void} whilst function \lstinline{f()} is {\em explicitly} declared as returning \lstinline{void}.  Since \lstinline{void} indicates the {\em absence} of a return value, neither \lstinline{g()} nor \lstinline{f()} are permitted to return a value.

\section{Types}
\label{c_err_types}

\subsection{``Subtype Error'' (401)}

\subsection{``Incomparable Operands'' (402)}

\subsection{``Record Type Required'' (403)}

\subsection{``Record Missing Field'' (404)}


\section{Statements}

Statements are used frequently in a Whiley program, and their syntax is defined in \S\ref{c_statements}.  The error messages reported in this section are those related to specific statement forms.  Other, more general, errors can also be reported for a statement (e.g. {\em type errors}, \S\ref{c_err_types}).

\subsection{``Invalid LVal'' (E501)}

An {\em invalid lval} error occurs when an invalid expression is used on the left-hand side of an assignment.  Only expressions which are also lval's maybe used in such a situation  (see~\S\ref{c_stmts_lval}).

\paragraph{Example.}  The following illustrates two invalid lval's:

\begin{lstlisting}
function f(int x):
   1 = x   // constant not valid lval
   x+1 = x // arithmetic expression not valid    
\end{lstlisting}

The first assignment statement is invalid because one cannot assign to a constant.  The second is invalid because one cannot assign to an arithmetic expression.

\subsection{``Invalid Destructuring LVal'' (E502)}

An {\em invalid destructuring lval} occurs when a destructuring assignment is used on the left-hand side, but the right-hand side returns an incorrect number of values.

\paragraph{Example.}  The following illustrates an invalid destructuring LVal:

\begin{lstlisting}
function f(int x, int y) -> int:
    return x+y

function g(int x, int y) -> int:
    x,y = f(x,y)
    return x - y
\end{lstlisting}

Here, the invocation of \lstinline{f()} in \lstinline{g()} uses an invalid destructuring assignment because \lstinline{f()} returns one value, the assignment expects two.

\subsection{``Variable Already Defined'' (E503)}

A {\em variable redefinition} occurs when a variable is declared with a name matching another variable already in scope.  This is an error because it is not permitted for one variable to shadow another.

\paragraph{Example.}  The following illustrates an example of a variable redefinition:

\begin{lstlisting}
function sum(int[] items) => int[]:
    int i = 0
    int r = 0
    //
    while i < |items|:
       //
       int r = items[i]
       i = i + 1
    //
    return r
\end{lstlisting}

Here, the \lstinline{while} loop attempts to declare a variable \lstinline{r}, but another variable \lstinline{r} was already declared beforehand.

\subsection{``Unreachable Code'' (E504)}

In a function or method, {\em unreachable code} arises when no possible execution path could reach them.  

\paragraph{Example.} The following illustrates some unreachable code:

\begin{lstlisting}
function abs(int x) -> int:
    //
    if x < 0:
        return -x
    else:
        return x
    //
    return 0 // unreachable
\end{lstlisting}

Here, the final \lstinline{return} statement can never be reached by any execution path through the \lstinline{abs()} function.  This is considered an error because it indicates something undesirable which may not have been intended.

\subsection{``Missing Return Value'' (E505)}

A {\em missing return value} occurs when a \lstinline{return} statement which does not return a value occurs within a function or method declared to return one or more values.  

\paragraph{Example.} The following illustrates a missing return value:

\begin{lstlisting}
function abs(int x) -> int:
    //
    if x < 0:
        return -x
    //
    return
\end{lstlisting}

The last \lstinline{return} statement in this function does not return a value, but the function is declared to return an \lstinline{int}.

\subsection{``Branch Always Taken'' (E506)}

A {\em branch always taken} occurs when a conditional branch is determine to always evaluate to either \lstinline{true} or \lstinline{false}.  

\paragraph{Example.} The following illustrates a branch always taken:

\begin{lstlisting}
function f(int x) -> int:
   if x is int:
       return x
   else:
       return -1
\end{lstlisting}

Here, the condition \lstinline{x is int} always evaluates to \lstinline{true} and, hence, the true branch of this conditional is always taken, whilst the false branch is never taken.

\subsection{``Break Outside of Loop'' (E507)}

A {\em break outside loop} occurs when a \lstinline{break} statement is given which is not contained within one or more loops.  This is an error because the break statement is used specifically to exit a loop early, and must be contained within the loop to be exited.  

\paragraph{Example.}  The following illustrates a break outside of a loop:

\begin{lstlisting}
function f(int x) -> int:
   break
   return x
\end{lstlisting}

Here, the \lstinline{break} statement is meaningless as it is not associated with a loop.

\subsection{``Duplicate Default Label'' (E508)}
A {\em duplicate default label} occurs when a \lstinline{switch} statement includes more than one \lstinline{default} label.  This is an error because at most one \lstinline{default} is permitted.

\paragraph{Example.}  The following illustrates an example of a duplicate \lstinline{default} label:

\begin{lstlisting}
function f(int x):
    switch x:
        case 0:
            return 0
        default:
            return 1
        default:
            return 2
\end{lstlisting}

Here, the \lstinline{switch} statement has two \lstinline{default} labels.  This must be an error as, otherwise, it would be ambiguous as to which executed.

\subsection{``Duplicate Case Label'' (E509)}
A {\em duplicate case label} occurs when a \lstinline{switch} statement includes more than one \lstinline{case} label matching the same value.  This is an error because at most one \lstinline{case} matching a given value is permitted.

\paragraph{Example.}  The following illustrates an example of a duplicate \lstinline{case} label:

\begin{lstlisting}
function f(int x):
    switch x:
        case 0:
            return 0
        case 0,1:
            return 1
        default:
            return 2
\end{lstlisting}

Here, the \lstinline{switch} statement has two \lstinline{case} labels, both of which match the value \lstinline{0}.  This must be an error as, otherwise, it would be ambiguous as to which executed.


\section{Expressions}

%\subsection{``Invalid File Access'' (000)}

%\subsection{``Invalid Package Access'' (000)}

%\subsection{``Resolution Error'' (000)}

Expressions typically form the bulk of a Whiley program, and their syntax is defined in \S\ref{c_expressions}.  The error messages reported in this section are those related to specific expression forms.  Other, more general, errors can also be reported for an expression (e.g. {\em type errors}, \S\ref{c_err_types}).

\subsection{``Variable Possibly Uninitialised'' (E601)}
\label{c_err_var_uninitialised}
A {\em variable possibly uninitialised} occurs when a variable may be used without being defined.  That is, when a simple path exists through the control-flow graph of a function or method from that variable's declaration to a use which contains no definition for that variable.  This error is reported as part of the {\em definite assignment} checking performed during compilation (see~\S\ref{c_definite_assignment}).

\paragraph{Example.}  The following illustrates a variable which is possibly uninitialised:

\begin{lstlisting}
function f(int x) => int:
   int y
   return x + y
\end{lstlisting}

Here, variable \lstinline{y} is definitely uninitialised in the expression ``\lstinline{x + y}''.  For more examples of variables which are possibly uninitialised, see~\S\ref{c_definite_assignment}.

\subsection{``Unknown Variable'' (E602)}

An {\em unknown variable} occurs when an attempt is made to access a variable which has not been declared in the current scope.  All variables must be declared before they can be used.

\paragraph{Example.}  The following illustrates an unknown variable:

\begin{lstlisting}
function f(int x) -> int:
   return x+y
\end{lstlisting}

Here, the \lstinline{return} statements refers to an unknown variable \lstinline{y}.  In contrast, the reference to variable \lstinline{x} is valid because \lstinline{x} has been declared within scope.

\subsection{``Unknown Function or Method'' (E603)}

An {\em unknown function or method} occurs when an attempt is made to access a function or method which is not visible in the current scope.  Functions and methods which are not declared in the same file as the invocation can be brought into the current scope using \lstinline{import} statements.  
\paragraph{Example.}  The following illustrates an invocation of an unknown function or method:

\begin{lstlisting}
function f(int x) -> int:
   return g(x)
\end{lstlisting}

Here, function \lstinline{g()} is not defined in the current source file and has not been brought into scope through an \lstinline{import} statement.

\subsection{``Ambiguous Coercion'' (E604)}

An {\em ambiguous coercion} occurs when the target of a cast expression is uncertain.  That is, when attempting to cast a value to a given type \lstinline{T}, but there is more than one way this can be achieved.  This error is reported as part of the {\em coercion check} performed during compilation.

\paragraph{Example.}  The following illustrates an ambiguous coercion:

\begin{lstlisting}
type Ambiguous is { int f1, real f2} | { real f1, int f2 }

function f(int x, int y) -> Ambiguous:
   return (Ammbiguous) {f1: x, f2: y}
\end{lstlisting}

The cast is ambiguous here because it's unclear whether, for example, \lstinline|{f1: 1, f2: 2}| should become \lstinline|{f1: 1.0, f2: 2}| or \lstinline|{f1: 1, f2: 2.0}|.

