\newcommand*{\glossfirstformat}[1]{\textit{#1}}
\usepackage[toc,xindy]{glossaries}
\makeglossary
\usepackage[xindy]{imakeidx}
\makeindex
\renewcommand{\glsdisplayfirst}[4]{\glossfirstformat{#1#4}}

% ===============================================
% A
% ===============================================

% ===============================================
% B
% ===============================================

\newglossaryentry{boolean_expression}{
  name={boolean expression},
  description={An \gls{expression} which evaluates to a value of type \lstinline{bool}}
}

% ===============================================
% C
% ===============================================

\newglossaryentry{compilation_unit}{
  name={compilation unit},
  description={A single unit of compilation.  In Whiley, this includes
    \gls{source_file}s and also binary \gls{wyil_file}s}
}

% ===============================================
% D
% ===============================================

% ===============================================
% E
% ===============================================

\newglossaryentry{expression}{
  name={expression},
  description={A combination of constants, variables and operators that, when evaluated, produce a single value.  Expressions in certain circumstances may have side effects}
}

% ===============================================
% F
% ===============================================

% ===============================================
% G
% ===============================================

% ===============================================
% H
% ===============================================

% ===============================================
% I
% ===============================================

% ===============================================
% J
% ===============================================

% ===============================================
% K
% ===============================================

% ===============================================
% L
% ===============================================

\newglossaryentry{loop_invariant}{
  name={loop invariant},
  description={A \gls{boolean_expression} which must hold on every iteration of a loop}
}

% ===============================================
% M
% ===============================================

% ===============================================
% N
% ===============================================

% ===============================================
% O
% ===============================================

% ===============================================
% P
% ===============================================

\newglossaryentry{package}{
  name={package},
  description={A unit of hierarchical organisation within the Whiley namespace.}
}

\newglossaryentry{precondition}{
  name={precondition},
  description={A logical condition over the parameters of a function
    or method which must be true immediately prior to execution of
    that function or method.}
}

\newglossaryentry{postcondition}{
  name={postcondition},
  description={A logical condition over the parameters and returns of
    a function or method which must be true immediately after
    execution of that function or method.}  }

% ===============================================
% Q
% ===============================================

% ===============================================
% R
% ===============================================

% ===============================================
% S
% ===============================================

\newglossaryentry{safety_critical_system}{
  name={safety critical system},
  description={A system which operates in a high-risk setting where failure can lead to loss of life, injury, significant damage or environmental harm}
}

\newglossaryentry{source_file}{
  name={source file},
  description={A file in which source code is located.  Source files
    for the Whiley programming language have the extension
    \lstinline{.whiley}.  In Whiley, source files must be compiled into a
    binary form before they can be executed.}
}

% ===============================================
% T
% ===============================================

\newglossaryentry{type}{
  name={type},
  description={An descriptor for a set of values, typically used to determine the set of values a given variable or \gls{expression} may hold}
}

% ===============================================
% U
% ===============================================

% ===============================================
% V
% ===============================================

\newglossaryentry{variable_declaration}{
  name={variable declaration},
  description={A statement which declares one or more variable(s) for use in a given scope.  Each variable is given a \gls{type} which limits the possible values it may hold, and may not already be declared in an enclosing scope}
}

\newglossaryentry{variable_initialiser}{
  name={variable initialiser},
  description={An optional \gls{expression} used to initialise variable(s) declared as part of a \gls{variable_declaration}}
}

\newglossaryentry{verifying_compiler}{
  name={verifying compiler},
  description={A compilers which employs automated mathematical and logical reasoning to check the correctness of the programs that it compiles}
}

% ===============================================
% W
% ===============================================


\newglossaryentry{wyil_file}{
  name={WyIL file},
  description={A compiled (i.e. binary) form of a Whiley \gls{source_file}}
}

% ===============================================
% X
% ===============================================

% ===============================================
% Y
% ===============================================

% ===============================================
% Z
% ===============================================

