\chapter{Introduction}

This document provides a specification of the {\em Whiley Programming Language}.  Whiley is a hybrid imperative and functional programming language designed to produce programs with as few errors as possible.  Whiley allows explicit specifications to be given for functions, methods and data structures, and employs a \gls{verifying_compiler} to check whether programs meet their specifications.  As such, Whiley is ideally suited for use in \gls{safety_critical_system}s.  However, there are many benefits to be gained from using Whiley in a general setting (e.g. improved documentation, maintainability, reliability, etc).  Finally, this document is {\em not} intended as a general introduction to the language, and the reader is referred to alternative documents for learning the language~\cite{Pearce14a}.


\section{Background}

Reliability of large software systems is a difficult problem facing
software engineering, where subtle errors can have disastrous
consequences.  Infamous examples include: the Therac-25 disaster where
a computer-operated X-ray machine gave lethal doses to
patients~\cite{LT93}; the 1988 worm which reeked havoc on the internet
by exploiting a buffer overrun~\cite{ER89}; the 1991 Patriot missile
failure where a rounding error resulted in the missile catastrophically
hitting a barracks~\cite{GAO}; and, the Ariane 5 rocket which exploded
shortly after launch because of an integer overflow, costing the ESA
an estimated \$500 million~\cite{ARIAN5}.

% The most widely used and accepted approach to improving software
% reliability is through extensive testing and manual code inspection.
% Whilst this does increase confidence, it cannot guarantee the absence
% of errors --- which is particularly problematic in a safety-critical
% setting.  Another successful approach is to prove the correctness of
% {\em models of software}, rather than of the software itself.  For
% example, model checkers
% (e.g~\cite{Clarke99,Holz01,BR02}) and SAT solvers % Mcm93,,CDHLPZ00
% (e.g.~\cite{MMZZM01,MFM04}) have proved highly % ,HDZM04,LS04,VB03
% effective at checking correctness properties of finite models of
% software systems, including microprocessor
% designs~\cite{VB03,Schub03}, flight-control
% systems~\cite{GH02,Choi05}, network protocols~\cite{Boli98,AC03} and
% spaceflight-control systems~\cite{HP00b}.  Some model checkers % ,NP02
% (e.g. CBMC~\cite{CKL04}, Java Pathfinder~\cite{HP00b},
% BLAST~\cite{HJMS03}, SLAM~\cite{BMMR01}) can also be applied directly
% on the program code although, in such cases, either significant
% abstraction is performed (hence, reducing the scope) or scalability is
% sacrificed.

Around 2003, Hoare proposed the creation of a {\em verifying compiler} as a grand challenge for computer science \cite{Hoare03}.  A verifying compiler ``{\em uses automated mathematical and logical reasoning to check the correctness of the programs that it compiles.}''  There have been numerous attempts to construct a verifying compiler system, although none has yet made it into the mainstream.  Early examples include that of King~\cite{King69}, Deutsch~\cite{Deutsch73}, the Gypsy Verification Environment~\cite{Good85} and the Stanford Pascal Verifier~\cite{LGHKMOPS95}.  More recently, the Extended Static Checker for Modula-3~\cite{DLNS98} which became the Extended Static Checker for Java (ESC/Java) --- a widely acclaimed and influential work~\cite{FLLNSS02}.  Building on this success was JML and its associated tooling which provided a standard notation for specifying functions in Java~\cite{LCCRC05}.  Finally, Microsoft developed the Spec\# system which is built on top of C\#~\cite{BLS04}.% ,BDFLS04,BCDJL06

\section{Goals}

The Whiley Programming Language has been designed from scratch in conjunction with a verifying compiler.  The intention is to provide an open framework for research in automated software verification.  The initial goal is to automatically eliminate common errors, such as {\em null dereferences}, {\em array-out-of-bounds}, {\em divide-by-zero} and more.  In the future, the intention is to consider more complex issues, such as termination, proof-carrying code and user-supplied proofs.

\section{History}

Development of the Whiley programming language begun in 2009 by Dr. David J. Pearce, at the time a lecturer in Computer Science at Victoria University of Wellington.  The accompanying website \url{http://whiley.org} went live in 2010, making the first versions of Whiley available for download.  Since then, Whiley has been in constant development with the majority of contributions being made by the original author.  Several scientific papers have published on different aspects of the language, including:

\begin{itemize}
\item {\bf Implementing a Language with Flow-Sensitive and Structural Typing on the JVM}.  David J. Pearce and James Noble. In {\em Proceedings of the Workshop on Bytecode Semantics, Verification, Analysis and Transformation (BYTECODE)}, 2011. 

\item {\bf Sound and Complete Flow Typing with Unions, Intersections and Negations}, David J. Pearce.  In {\em Proceedings of the Conference on Verification, Model Checking and Abstract Interpretation (VMCAI)}, pages 335--354, 2013

\item {\bf A Calculus for Constraint-Based Flow Typing}. David J. Pearce. In {\em Proceedings of the Workshop on Formal Techniques for Java-like Languages (FTFJP)}, Article 7, 2013.

\item {\bf Whiley: a Platform for Research in Software Verification}. David J. Pearce and Lindsay Groves. In {\em Proceedings of the Conference on Software Language Engineering (SLE)}, pages 238-–248, 2013

\item {\bf Reflections on Verifying Software with Whiley}. David J. Pearce and Lindsay Groves. In {\em Proceedings of the Workshop on Formal Techniques for Safety-Critical Software (FTSCS)}, 2013

\end{itemize}